Simulation of Nano Particles in a Laser Trap

### 2017-02-28 doc/draft4.tex
more writing

\begin{document}
\title{Simulation of Nano Particles in a Laser Trap}
\author{Mathias H\"old, BSc.}
\date{2016}
\maketitle
\thispagestyle{empty}
\newpage
\tableofcontents
\newpage
\section{Introduction}
%something about computer science in general, optical tweezers and the need to study the
%systems on computers yada yada
In the year 1953, Metropolis et al.~\cite{Metropolis1953} used the MANIAC (Mathematical Analyzer Numerical Integrator And Computer) 
at the Los Alamos National Laboratories in the United States to
simulate particles in a liquid state to calculate the equation of state. This was the first publication in which scientists used a computer to
calculate numerical results based on a physical theory in a process that is to this day called \textit{Metropolis Monte Carlo simulation}. 
To say that since then a lot has
changed in the field of computer simulation (or computers in general) would be a vast understatement.\\
Not only has the processing power of computers increased exponentially (Moore's law), which allows the simulation of a greater number of particles for
a longer period of time, but also new techniques have been developed, such as molecular dynamics. These techniques are not only used by physicists,
but also find applications in the fields of chemistry, biology, etc., making it possible to apply them to virtually anything. In thhis thesis we 
apply methods of simulation to a field of active research: optical trapping.
Optical trapping is a technique that has been used and developed for over 40 years~\cite{Marago2013}. The idea for this goes back to the beginning of
the 20th century, where Lebedev~\cite{Lebedew1901}, Nichols and Hull~\cite{Nichols1901} demonstrated the existence of a force that light exerts on 
matter. Due to the technical limitations at that time, the idea was discarded until 1970, where Ashkin~\cite{Ashkin2000} used lasers to manipulate the
motion of neutral atoms and micrometer sized objects and thus brought the idea of optical trapping back to life.\\
With this advancement it was possible to develop methods for trapping and manipulating objects of different sizes ranging from the micrometer scale
down to the subnano scale. At first, this technique was used to either trap objects in the micrometer range (like cells) or in the subnano scale,
where lasers were used for the cooling of atoms. The range between those two, with objects at the scale of one to a hundred nanometer, was not
explored as much because neither downscaling the methods for micrometer sized objects nor upscaling the methods for atoms seemed feasible. In recent
years however, new developments in the field opened the possibilities to work with a variety of objects in the desired nanometer scale, such as metal
nanoparticles, nano tubes, quantum dots and much more. This non-invasive method of trapping objects allows very precise measurements of various
properties and also allows the cooling of nanoparticles towards their quantum mechanical ground state.\\
The small size of the objects used in the optical tweezer experiments make them amenable candidates for computer simulations. In this thesis, I will
model an optical tweezer setup, based on a real experiment, to investigate different temperatures in a non-equilibrium state.\\
The first chapter will be the introduction of the experiment and the introduction of the problem statement. The second chapter contains basic
information on the techniques used to model the experiment and simulate the dynamic properties. In the third chapter, I will present my findings which
are followed by conclusions and further remarks.

@@

On the left hand side we have the position $x$ (and its derivatives $\dot{x}$, the velocity and $\ddot{x}$, the acceleration), the friction 
coefficient $\Gamma_0$ and the angular frequency $\Omega_0$ that describes the fluctuation along the chosen axis. On

@@

@@

interaction $T_\text{em}$ (for emerging), the surface temperature of the nano particle $T_\text{sur}$ and the temperature of the 
center of mass of the nano particle $T_\text{COM}$. The experimental setup

@@

Since the two temperatures of the gas, $T_\text{imp}$ and $T_\text{em}$, are not equal, the gas surrounding the nano particle is not in a thermal
equilibrium, which is in contrast to the assumptions made in many optical tweezer experiments. Since the gas particles only interact with the nano particle

@@

\newpage
\section{Simulation}
%simulation techniques and general concept of the simulation itself, used methods and so on 
The problem at hand can be studied on an atomic level with the use of computer simulation. There is a variety of methods for computer simulations
that are widely used, one of which being molecular dynamics (MD) simulations.\\ 
The goal is to simulate the experiment depicted in Fig.~\ref{fig:setup} as accurately as possible. To achieve this, the setup has to be broken down
into individual pieces that can be modeled by using existing methods. The nano particle will be approximated by a cube of particles, located on the
sites of a FCC
(face-centered cubic) lattice that interact via a Lennard-Jones potential. The laser will be broken down into its two main purposes, 
as it acts as a trapping device as well as a heat source for the nano particle. Surrounding the nano particle will be a gas chamber, 
that acts as a thermostat in the simulation, which will be modeled as a cubic box surrounding the nano particle. 
This box is filled with gas particles that interact with the nano particle via a soft-sphere potential and do not interact with one another.\\
The following section will give a brief overview of the concepts used to simulate every part of the experiment.

@@

\subsection{The Laser Beam - Trapping}
As mentioned above, the glass particle is trapped in the laser beam and the position of the particle is localized. This behaviour
has to be modeled as well.\\
The approximation of the original paper \cite{Gieseler2014}, where the movements in the three spatial directions are decoupled will be
used in the model as well. As to the model of the trapping force itself, the most straightforward approach is to use a harmonic oscillator potential.
The force and the corresponding potential can be written as
\begin{eqnarray}
    \mathbf{F} &=& -k \Big[\mathbf{x}-\mathbf{x}_0\Big]\\
    U &=& \frac12 k \Big[\mathbf{x}-\mathbf{x}_0\Big]^2
\end{eqnarray}
where $\mathbf{x}_0$ is the position of minimal potential energy. 
Since the trap is acting on the whole system, the force is acting on the center of mass. 
To calculate the center of mass positions and velocities, the positions and velocities of all particles have to be summed up and 
divided by the particle number:
\begin{equation}
    \mathbf{r}_\text{COM} = \frac1N\sum_{i=1}^N \mathbf{r}_i
\end{equation}

\subsection{Surrounding Gas - Thermostat}
Without any equilibrating mechanism the setup described by now would lead to the system heating up indefinitely, which is not desirable. Furthermore,
the goal is to recreate the experiment as close as possible. Since the nano particle is trapped in a laser beam within a vacuum chamber, the
surrounding gas has to be modeled as well. In the next step we will introduce the surrounding gas of the gas chamber that will absorb some of the 
energy in the system, leading to the final state.\\
Generally, pressure is introduced to the system by surrounding the object of interest (in this case the glass nano sphere) with a thermostatting
pressure medium.
There are two main requirements for the choice of such a pressure medium: the exerted pressure must be hydrostatic and the computation of the
interaction between the pressure medium and the object of interest must not take up a lot of resources.\\
The model used in this thesis was developed by Gr\"unwald and Dellago \cite{Gruenwald2006} and uses an ideal gas of non-interacting particles as
thermostatting pressure medium. The particles of this pressure medium flow into the simulation from an outside volume, whose geometry is based on the form of the
object of interest (this will be referred to as the minimal volume of cells) and leave the simulation box if their position reaches the boundary of
this minimal volume of cells. This behaviour is very close to the real experiment which makes this thermostat an ideal candidate for the simulation of
the experiment. 

@@

\newpage
\section{Results}
% results with different parameters, graphs, etc.
Since the simulation consists of various parts that have to work together, we first need to check if every part itself works.\\
\subsection{The Crystal}
The first element of the simulation is the setup of the FCC lattice on which the particles are placed on. To check if this is done correctly, we can
use a visualisation and rendering software such as VMD\cite{Humphrey1996} or OVITO\cite{Stukowski2010}. The rendered image of the nano particle 
can be seen in Fig.~\ref{fig:glass}. The chosen perspective shows the structure of the FCC lattice, where 3 layers are forming 
where the atoms take the position on the gap between the atoms from the layer below.\\
\begin{figure}[h]
    \begin{center}
        \includegraphics[scale=0.2]{images/fcc_rendering_layer_newcoloring.png}
        \caption{A rendered image of the FCC lattice. The chosen color-coding along the (1,1,1) Miller-index 
                highlights the layers for this perspective. The perspective is
                chosen to see the layering of the FCC lattice with its 3 layer structure. The black box acts as a reference point to outline the 
                geometry of the system.}
            %Rendering of the model of the nano crystal before starting the simulation. 
            %The chosen perspective shows the 3 layers of the FCC crystal with a
                    %These layers consist of atoms
                    %taking the position above the gap between two of the atoms from the layer below and match the positions of the atoms 2 layers
                %below. The black box is the outline of the particles' positions and shows the geometry of the system.}
        \label{fig:glass}
    \end{center}
\end{figure}

@@

The velocity Verlet part of the simulation is taking place in a NVE ensemble from a thermodynamic point of view. 

>>>>

\begin{document}
\title{Simulation of Nano Particles in a Laser Trap}
\author{Mathias H\"old, BSc.}
\date{2016}
\maketitle
\thispagestyle{empty}
\tableofcontents
\section{Introduction}
%something about computer science in general, optical tweezers and the need to study the
%systems on computers yada yada
In the year 1953, Metropolis et al.~\cite{Metropolis1953} used the MANIAC (Mathematical Analyzer Numerical Integrator And Computer) 
at the Los Alamos National Laboratories in the United States to
simulate particles in a liquid state to calculate the equation of state. This was the first publication in which scientists used a computer to
calculate numerical results based on a physical theory in a process that is to this day called \textit{Metropolis Monte Carlo simulation}. 
To say that since then a lot has
changed in the field of computer simulation (or computers in general) would be a vast understatement.\\
Not only has the processing power of computers increased exponentially (Moore's law), which allows the simulation of a greater number of particles for
a longer period of time, but also new techniques have been developed, such as molecular dynamics, transition path sampling, finite element method etc. 
These techniques are not only used by physicists,
but also find applications in the fields of chemistry, biology, etc., making it possible to apply them to virtually anything. In this thesis we 
apply methods of simulation to a field of active research: optical trapping. 
Optical trapping refers to the experimental method of localising small objects with a laser beam. The laser is highly focused on one point, where 
the object is trapped due to the gradient force of the laser. The size of the trapped objects ranges from a few micrometers down to the subatomic 
scale.\\
%Optical trapping refers to experiments where a focused laser beam is 
%used to keep small objects with a size in the range of a few micrometers down to a subatomic scahe in place.\\
Optical trapping is a technique that has been used and developed for over 40 years~\cite{Marago2013}. The idea for this goes back to the beginning of
the 20th century, when Lebedev~\cite{Lebedew1901}, Nichols and Hull~\cite{Nichols1901} demonstrated the existence of a force that light exerts on 
matter. Due to the technical limitations at that time, the idea was discarded until 1970, where Ashkin~\cite{Ashkin2000} used lasers to manipulate the
motion of neutral atoms and micrometer sized objects and thus brought the idea of optical trapping back to life.\\
With this advancement it was possible to develop methods for trapping and manipulating objects of different sizes ranging from the micrometer scale
down to the subnano scale. At first, this technique was used to either trap objects in the micrometer range (like cells) or in the subnano scale,
where lasers were used for the cooling of atoms. The range between those two, with objects at the scale of one to a hundred nanometer, was not
explored as much because neither downscaling the methods for micrometer sized objects nor upscaling the methods for atoms seemed feasible. In recent
years however, new developments in the field opened the possibilities to work with a variety of objects in the desired nanometer scale, such as metal
nanoparticles~\cite{Svoboda1994}, nano tubes~\cite{Tan2004}, quantum dots~\cite{Jauffred2008} and much more. 
This non-invasive method of trapping objects allows very precise measurements of various
properties and also allows the cooling of nanoparticles towards their quantum mechanical ground state.\\
The small size of the objects used in the optical tweezer experiments make them amenable candidates for computer simulations. In this thesis, I will
model an optical tweezer setup, based on a real experiment, to investigate different temperatures in a non-equilibrium state. 
The experiment features a silica nano sphere trapped in a laser beam within a vacuum chamber. To simulate this experiment, each of these individual parts 
of the experiment need to be modeled. The silica nano sphere is approximated by a system of argon atoms that interact with each other via a 
Lennard-Jones potential. These atoms are absorbing energy from the laser and are thus heated up using the enhanced heat exchange algorithm (eHEX) 
and a harmonic trap is keeping them in place. The gas particles in the gas chamber are acting as a thermostat/barostat and are modeled by a ideal
gas pressure bath.\\
In the first chapter, I will introduce two optical tweezer experiments: one conducted by Gieseler et al. to investigate the relaxation of a nano particle
from a non-equilibrium steady state and another conducted by Millen et al. to examine different temperature regimes in an optical tweezer experiment. 
This is followed by the problem statement. The second chapter contains basic
information on the techniques used to model the individual parts of the experiment and 
simulate the dynamic properties. In the third chapter, I will present my findings which
are followed by conclusions and further remarks.

@@

On the left hand side of this equation we have the position $x$ of the nano particle (and its derivatives $\dot{x}$, the velocity and $\ddot{x}$, 
the acceleration), the friction coefficient $\Gamma_0$ and the angular frequency $\Omega_0$ that describes the fluctuation along the chosen axis. On

@@

The other force, $F_\text{ext}$ is a time-dependent external force that is applied to the nano particle to create a non-equilibrium steady state. At
time $t = t_\text{off}$ the force is deactivated and the particle relaxes to a thermally equilibrated state in the laser trap. The evolution of the nano 
particle from the non equilibrium steady state to the equilibrated state is examined to investigate the fluctuation theorem, which can be written 
down as 
    \label{eq:fluctuationtheorem}
    \frac{p(-\Delta S)}{\Delta S} = e^{-\Delta S}
where $\Delta S$ is the relative entropy change given by
    \Delta S = \beta_0 Q + \Delta \phi
In the above equation, $Q$ is the heat absorbed by the bath, $\beta_0$ is the reciprocal temperature at which the heat is absorbed and 
$\Delta \phi$ is the difference of the trajectory dependent entropy. The fluctuation theorem~\cite{Searles1999}, as expressed in 
\eqref{eq:fluctuationtheorem}, states that there is a measurable probablity for a finite system to violate the second law of thermodynamics
for finite times. This does not stand in contradiction to the second law of thermodynamics itself, as it applies only to short times and does not 
happen for large systems or long time scales.\\
The investigation of the fluctuation theorem is performed for two different external forces. The first one is a feedback force and the second 
one is the feedback force combined with a parametric drive. The results show that the relaxation process for both initial non-equilibrium steady 
states generated by the external forces demonstrates the validity of the fluctuation theorem.\\

@@

interaction $T_\text{em}$ (for \textit{emerging}), the \textit{surface} temperature of the nano particle $T_\text{sur}$ and the temperature of the 
\textit{center of mass} of the nano particle $T_\text{COM}$. The experimental setup

@@

The nano particle in the trap is absorbing heat from the laser which causes the surface temperature of the nano particle to be higher than the 
temperature of the surrounding gas. The impinging gas particles interact with the sphere, absorb some of the energy and emerge with a different
temperature, which is in general not equal to the surface temperature. This means that
the two temperatures of the gas, $T_\text{imp}$ and $T_\text{em}$, are not equal, the gas surrounding the nano particle is not in a thermal
equilibrium, which is in contrast to the assumptions made in many optical tweezer experiments. 
Since the gas particles only interact with the nano particle

@@

\newpage
\section{Simulation}
%simulation techniques and general concept of the simulation itself, used methods and so on 
The problem at hand can be studied on an atomic level with the use of computer simulation. There is a variety of methods for computer simulations
that are widely used, one of which being molecular dynamics (MD) simulations.\\ 
The goal is to simulate the experiment depicted in Fig.~\ref{fig:setup} as accurately as possible. To achieve this, the setup has to be broken down
into individual pieces that can be modeled by using existing methods. The nano particle will be approximated by a cube of particles, located on the
sites of a FCC
(face-centered cubic) lattice that interact via a Lennard-Jones potential. The laser serves two purposes: heating up and trapping the particle. The 
nano particle will be heated up using the eHEX algorithm and trapped using a harmonic potential.
%Since the laser serves a dual purpose -- heating and trapping --
%both of these are modeled separately.
%The laser will be broken down into its two main purposes, 
%as it acts as a trapping device as well as a heat source for the nano particle. 
Surrounding the nano particle will be a gas chamber, 
that acts as a thermostat in the simulation, which will be modeled as a cubic box surrounding the nano particle. 
This box is filled with gas particles that interact with the nano particle via a soft-sphere potential and do not interact with one another.\\
The following section will give a brief overview of the concepts used to simulate every part of the experiment.

@@

\subsection{The Laser Beam - Trapping}
As mentioned above, the glass particle is trapped in the laser beam and the position of the particle is localized. This behaviour
has to be modeled as well.\\
The approximation of the original paper \cite{Gieseler2014}, where the movements in the three spatial directions are decoupled will be
used in the model as well. As to the model of the trapping force itself, the most straightforward approach is to use a harmonic oscillator potential.
The force and the corresponding potential can be written as
\begin{eqnarray}
    \mathbf{F} &=& -k \Big[\mathbf{r}_\text{COM}-\mathbf{x}_0\Big]\\
    U &=& \frac12 k \Big[\mathbf{r}_\text{COM}-\mathbf{x}_0\Big]^2
\end{eqnarray}
where $\mathbf{x}_0$ is the position of minimal potential energy, $\mathbf{r}_\text{COM}$ 
is the position of the center of mass and $k$ is the spring constant.
%Since the trap is acting on the whole system, the force is acting on the center of mass. 
To calculate the center of mass positions and velocities, the positions and velocities of all particles have to be summed up and 
divided by the particle number:
\begin{equation}
    \mathbf{r}_\text{COM} = \frac1N\sum_{i=1}^N \mathbf{r}_i
\end{equation}
To use this in the simulation, the distance of the center of mass to the desired point of minimal potential energy and then the corresponding 
force is calculated and added to the interaction force of each atom.

\subsection{Surrounding Gas - Thermostat}
Without any equilibrating mechanism the setup described by now would lead to the system heating up indefinitely, which is not desirable. Furthermore,
the goal is to recreate the experiment as close as possible. Since the nano particle is trapped in a laser beam within a vacuum chamber, the
surrounding gas has to be modeled as well. In the next step we will introduce the surrounding gas of the gas chamber that will absorb some of the 
energy in the system, leading to the final state.\\
Generally, pressure is introduced to the system by surrounding the object of interest (in this case the glass nano sphere) with a thermostatting
pressure medium.
There are two main requirements for the choice of such a pressure medium: the exerted pressure must be hydrostatic and the computation of the
interaction between the pressure medium and the object of interest must not take up a lot of resources.\\
The model used in this thesis was developed by Gr\"unwald and Dellago \cite{Gruenwald2006} and uses an ideal gas of non-interacting particles as
thermostatting pressure medium. The particles of this pressure medium flow into the simulation from an outside volume, whose geometry is based on the form of the
object of interest (this will be referred to as the minimal volume of cells) and leave the simulation box if their position reaches the boundary of
this minimal volume of cells. The minimal volume of cells is based on cell lists used in the simulation. Cell lists are a method of saving 
computing time by dividing the simulation box in cells with a side length equal to or greater than the cutoff radius $r_C$. The particles are then
sorted into these cells and the interaction between the particles is only calculated for other particles within the same and neighbouring cells. 
For the minimum volume of cells used in the thermostat algorithm all cells filled with particles and one additional cell for the interaction partners
is used.\\
This algorithm is a very good approximation of the situation in the experiment.  There the gas particles only interact with the nano 
particle in the laser trap and not with eatch other. After this interaction the
gas particles emerge with a slightly higher temperature and leave the vicinity of the nano particle. Moreover, the algorithm uses an ideal gas
as a pressure medium, which allows us to simulate the dynamics of the interaction that is happening in the experiment.\\

%This behaviour is very close to the real experiment which makes this thermostat an ideal candidate for the simulation of
%the experiment. There the gas particles only interact with the nano particle in the laser trap and not with eatch other. After this interaction the
%gas particles emerge with a slightly higher temperature and leave the vicinity of the nano particle. 

@@

\newpage
\section{Results}
% results with different parameters, graphs, etc.
Since the simulation consists of various parts that have to work together, we first need to check if every part itself works.\\
\subsection{The Crystal}
The first element of the simulation is the setup of the FCC lattice on which the particles are placed on. To check if this is done correctly, we can
use a visualisation and rendering software such as VMD\cite{Humphrey1996} or OVITO\cite{Stukowski2010}. The rendered image of the nano particle 
can be seen in Fig.~\ref{fig:glass}. To see that the particles form a FCC lattice a perspective is chosen, where the three layers of the FCC lattice
are visible. \\
These layers are generated by stacking the atoms onto the gap between two atoms. In a FCC lattice there are three distinct layers, which means that
the atoms of the fourth layer are on the same position as the atoms in the first layer.
%The chosen perspective shows the structure of the FCC lattice, where 3 layers are forming 
%where the atoms take the position on the gap between the atoms from the layer below.\\
\begin{figure}[h]
    \begin{center}
        \includegraphics[scale=0.2]{images/fcc_rendering_layer_newcoloring.png}
        \caption{A rendered image of the FCC lattice. The chosen color-coding along the (1,1,1) Miller-index 
                highlights the layers for this perspective. The perspective is
                chosen to see the layering of the FCC lattice with its 3 layer structure. The black box acts as a reference point to outline the 
                geometry of the system.}
            %Rendering of the model of the nano crystal before starting the simulation. 
            %The chosen perspective shows the 3 layers of the FCC crystal with a
                    %These layers consist of atoms
                    %taking the position above the gap between two of the atoms from the layer below and match the positions of the atoms 2 layers
                %below. The black box is the outline of the particles' positions and shows the geometry of the system.}
        \label{fig:glass}
    \end{center}
\end{figure}

@@

The simulation of the nano particle without the gas is taking place in a NVE ensemble from a thermodynamic point of view. 
